\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
\bibliographystyle{plainurl}
\title{Anders: Modal Homotopy Type System}
\titlerunning{CCHM/HTS}
\author{Namdak Tonpa}{Groupoid Infinity}{maxim@synrc.com}{https://orcid.org/0000-0001-7127-8796}{}
\author{Siegmentation Fault}{Groupoid Infinity}{siegmentationfault@yandex.ru}{}{}
\authorrunning{Namdak Tonpa and Siegmentation Fault} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
\Copyright{Namdak Tonpa and Siegmentation Fault} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003753.10003754.10003733</concept_id>
<concept_desc>Theory of computation~Lambda calculus</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[300]{Theory of computation~Lambda calculus}
\keywords{Homotopy Type System, Cubical Type Theory}

%\category{} %optional, e.g. invited paper

%\relatedversion{https://groupoid.space/anders.pdf} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional
%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

%\acknowledgements{Univalent people}%optional

\nolinenumbers %uncomment to disable line numbering
\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\EventEditors{John Q. Open and Joan R. Access}
%\EventNoEds{2}
%\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
%\EventShortTitle{CVIT 2016}
%\EventAcronym{CVIT}
%\EventYear{2016}
%\EventDate{December 24--27, 2016}
%\EventLocation{Little Whinging, United Kingdom}
%\EventLogo{}
%\SeriesVolume{42}
%\ArticleNo{23}

\frenchspacing

\newcommand{\tabstyle}[0]{\scriptsize\ttfamily\fontseries{l}\selectfont}

\begin{document}

\maketitle

\begin{abstract}
Here is presented a reincarnation of \textbf{cubicaltt} called \textbf{anders}.
\end{abstract}

\section{Introduction}
\label{sec:typesetting-summary}

\textbf{Anders} is a Modal HoTT proof assistant based on: classical MLTT-80 \cite{MLTT80}
with 0, 1, 2, W types; CCHM \cite{CCHM} in CHM flavour as cubical type system with
hcomp/transp Kan operations; HTS \cite{HTS} strict equality on pretypes; de Rham \cite{deRham}
stack modality primitives. We tend not to touch general recursive higher inductive schemes yet,
instead we will try to express as much HIT as possible through W, Coequlizer and HubSpokes
Disc in the style of HoTT/Coq homotopy library and Three-HIT theorem.

The HTS language proposed by Voevodsky exposes two different presheaf models of type theory:
the inner one is homotopy type system presheaf that models HoTT and the outer one is
traditional Martin-Löf type system presheaf that models set theory with UIP.
The motivation behind this doubling is to have an ability to express semisemplicial types.
Theoretical work on merging inner and outer languages was continued in 2LTT \cite{2LTT}.

$\mathbf{Installation}$. While we are on our road to Lean-like tactic language, currently we are at the stage of
regular cubical HTS type checker with CHM-style [5] primitives. You may try it from Github
sources: \footnote{\url{https://github.com/groupoid/anders}}{groupoid/anders} or install
through OPAM package manager. Main commands are \textbf{check} (to check a program)
and \textbf{repl} (to enter the proof shell).

\begin{table}[ht]
\centering
\begin{tabular}{l}
\$ opam install anders
\end{tabular}
\end{table}

Anders is fast, idiomatic and educational (think of optimized Mini-TT). We carefully draw the favourite Lean-compatible
syntax to fit 200 LOC in Menhir. The CHM kernel is 1K LOC. Whole Anders compiles under 2
seconds and checks all the base library under 1 second [i7-8700]. Anders proof assistant
as Homotopy Type System comes with its own Homotopy Library.

\section{Syntax}

The syntax resembles original syntax of the reference CCHM type checker cubicaltt,
is slightly compatible with Lean syntax and contains the full set of Cubical Agda \cite{CubicalAgda}
primitives (except generic higher inductive schemes).

Here is given the mathematical pseudo-code notation of the language
expressions that come immediately after parsing. The core syntax definition of HTS language
corresponds to the type defined in OCaml module:

\begin{table}[ht]
\centering
\begin{tabular}{rl}
 cosmos :=&$\textbf{U}_j \ |\ \textbf{V}_k$ \\
    var :=&$\textbf{var}\ name\ |\ \textbf{hole}$ \\
     pi :=&$\Pi\ name\ E\ E\ |\ \lambda\ name\ E\ E\ |\ E\ E$ \\
  sigma :=&$\Sigma\ name\ E\ E\ |\ (E,E)\ |\ E.1\ |\ E.2$ \\
      0 :=&$\textbf{0}\ |\ \textbf{ind$_0$}\ \textrm{E}\ \textrm{E}\ \textrm{E}$ \\
      1 :=&$\textbf{1}\ |\ \star\ |\ \textbf{ind$_1$}\ \textrm{E}\ \textrm{E}\ \textrm{E}$ \\
      2 :=&$\textbf{2}\ |\ 0_2\ |\ 1_2\ |\ \textbf{ind$_2$}\ \textrm{E}\ \textrm{E}\ \textrm{E}$ \\
      W :=&$\textbf{W}\ \textrm{ident}\ \textrm{E}\ \textrm{E}\ |\ \textbf{sup}\ \textrm{E}\ \textrm{E}\ |\ \textbf{ind$_W$}\ \textrm{E}\ \textrm{E}$ \\
     id :=&$\textbf{Id}\ E\ |\ \textbf{ref}\ E\ |\ \textbf{id$_J$}\ E$ \\
   path :=&$\textbf{Path}\ E\ |\ E^i\ |\ E\ @\ E$ \\
      I :=&$\textbf{I}\ |\ 0\ |\ 1\ |\ E\ \bigvee\ E\ |\ E\ \bigwedge\ E\ |\ \neg E$ \\
   part :=&$\textbf{Partial}\ E\ E\ |\ \textbf{[}\ (E=I) \rightarrow E, ...\ \textbf{]}$ \\
    sub :=&$\textbf{inc}\ E\ |\ \textbf{ouc}\ E\ |\ E\ \textbf{[}\ I\ \mapsto\ E\ \textbf{]}$ \\
    kan :=&$\textbf{transp}\ E\ E\ |\ \textbf{hcomp}\ E$ \\
   glue :=&$\textbf{Glue}\ E\ |\ \textbf{glue}\ E\ |\ \textbf{unglue}\ E\ E$ \\
     Im :=&$\textbf{Im}\ \mathrm{E}\ |\ \textbf{Inf}\ \mathrm{E}\ |\ \textbf{Join}\ \mathrm{E}\ |\ \textbf{ind$_{Im}$}\ \mathrm{E}\ \mathrm{E}$ \\
\end{tabular}
\end{table}

Further Menhir BNF notation will be used to describe the top-level language E parser.

\begin{table}[ht]
\centering
\begin{tabular}{rl}
      E :=&$\textrm{cosmos}\ |\ \textrm{var}\ |\ \textrm{MLTT}\ |\ \textrm{CCHM}\ |\ \textrm{Im}$ \\
   CCHM :=&$\textrm{path}\ |\ \textrm{I}\ |\ \textrm{part}\ |\ \textrm{sub}\ |\ \textrm{kan}\ |\ \textrm{glue}$ \\
   MLTT :=&$\textrm{pi}\ |\ \textrm{sigma}\ |\ \textrm{id}$ \\
\end{tabular}
\end{table}

\newpage

\textbf{Keywords}. The words of a top-level language (file or repl) consist of keywords or identifiers.
The keywords are following: \textbf{module}, \textbf{where}, \textbf{import}, \textbf{option}, \textbf{def}, \textbf{axiom},
\textbf{postulate}, \textbf{theorem}, \textbf{(}, \textbf{)}, \textbf{[}, \textbf{]}, \textbf{<}, \textbf{>},
\textbf{/}, \textbf{.1}, \textbf{.2}, \textbf{$\Pi$}, \textbf{$\Sigma$}, \textbf{,}, \textbf{$\lambda$}, \textbf{$\times$}, \textbf{$\rightarrow$}, \textbf{:}, \textbf{:=},
\textbf{$\mapsto$}, \textbf{U}, \textbf{V}, \textbf{$\bigvee$}, \textbf{$\bigwedge$}, \textbf{-}, \textbf{+}, \textbf{@}, \textbf{PathP},
\textbf{transp}, \textbf{hcomp}, \textbf{zero}, \textbf{one}, \textbf{Partial}, \textbf{inc},
\textbf{ouc}, \textbf{interval}, \textbf{inductive}, \textbf{Glue}, \textbf{glue}, \textbf{unglue}.\\

\textbf{Indentifiers}. Identifiers support UTF-8. Indentifiers couldn't start with $:$, $-$, $\rightarrow$. Sample identifiers:
$\neg$-of-$\bigvee$, 1$\rightarrow$1, is-?, =, \$$\sim$]!005x, $\infty$, x$\rightarrow$Nat.\\

\textbf{Modules}. Modules represent files with declarations. More accurate, BNF notation of module consists of imports, options and declarations.
\begin{table}[ht]
\tabstyle
\begin{tabular}{l}
\textbf{menhir} \\
\ \ \ start <Module.file> file \\
\ \ \ start <Module.command> repl \\
\ \ \ repl : COLON IDENT exp1 EOF | COLON IDENT EOF | exp0 EOF | EOF \\
\ \ \ file : MODULE IDENT WHERE line* EOF \\
\ \ \ path : IDENT \\
\ \ \ line : IMPORT path+ | OPTION IDENT IDENT | declarations \\
\end{tabular}
\end{table}

\textbf{Imports}. The import construction supports file folder structure (without file extensions) by using reserved symbol / for hierarchy walking.\\

\textbf{Options}. Each option holds bool value. Language supports following options: 1) girard (enables U : U); 2) pre-eval (normalization cache);
3) impredicative (infinite hierarchy with impredicativity rule);
In Anders you can enable or disable language core types, adjust syntaxes or tune inner variables of the type checker.
Here is the example how to setup minimal core able to prove internalization of \textbf{MLTT-73} variation (Path instead of Id and no inductive types, see base library):
In order to turn HIT into ordinary CiC calculus you may say:

\begin{table}[ht]
\tabstyle
\begin{tabular}{l}
\textbf{option} HIT false \\
\textbf{option} CCHM false \\
\textbf{option} MLTT true \\
\end{tabular}
\end{table}

\textbf{Declarations}. Language supports following top level declarations:
1) axiom (non-computable declaration that breakes normalization);
2) postulate (alternative or inverted axiom that can preserve consistency);
3) definition (almost any explicit term or type in type theory);
4) lemma (helper in big game);
5) theorem (something valuable or complex enough).


\begin{table}[ht!]
\tabstyle
\begin{tabular}{l}
\textbf{axiom} isProp (A : U) : U \\
\textbf{def} isSet (A : U) : U := \textbf{$\Pi$} (a b : A) (x y : Path A a b), Path (Path A a b) x y
\end{tabular}
\end{table}

Sample declarations. For example, signature isProp (A : U) of type U could be
defined as normalization-blocking axiom without proof-term or by providing proof-term as definition.

In this example (A : U), (a b : A) and (x y : Path A a b) are called telescopes.
Each telescope consists of a series of lenses or empty. Each lense provides a
set of variables of the same type. Telescope defines parameters of a declaration.
Types in a telescope, type of a declaration and a proof-terms are a language expressions exp1.

\begin{table}[ht]
\tabstyle
\begin{tabular}{l}
\textbf{menhir} \\
\ \ \ ident : IRREF | IDENT \\
\ \ \ lense : LPARENS ident+ COLON exp1 RPARENS \\
\ \ \ telescope : lense telescope \\
\ \ \ params : telescope | [] \\
\ \ \ declarations : \\
\ \ \ \ \ \ | DEF IDENT params DEFEQ exp1 \\
\ \ \ \ \ \ | DEF IDENT params COLON exp1 DEFEQ exp1 \\
\ \ \ \ \ \ | AXIOM IDENT params COLON exp1
\end{tabular}
\end{table}

\textbf{Expressions}. All atomic language expressions are grouped by four categories:
exp0 (pair constructions), exp1 (non neutral constructions), exp2 (path and pi applcations),
exp3 (neutral constructions).

\begin{table}[ht!]
\tabstyle
\begin{tabular}{ll}
\textbf{menhir} \\
\multicolumn{2}{l}{\ \ \ face : LPARENS IDENT IDENT IDENT RPARENS } \\
\multicolumn{2}{l}{\ \ \ part : face+ ARROW exp1 } \\
\multicolumn{2}{l}{\ \ \ exp0 : exp1 COMMA exp0 | exp1 } \\
\multicolumn{2}{l}{\ \ \ exp1 : LSQ separated(COMMA, part) RSQ } \\
\ \ \ \ \ \ | LAM telescope COMMA exp1   & | PI telescope COMMA exp1 \\
\ \ \ \ \ \ | SIGMA telescope COMMA exp1 & | LSQ IRREF ARROW exp1 RSQ \\
\ \ \ \ \ \ | LT ident+ GT exp1          & | exp2 ARROW exp1 \\
\ \ \ \ \ \ | exp2 PROD exp1             & | exp2 \\
\end{tabular}
\end{table}

The LR parsers demand to define exp1 as expressions that cannot be used (without a parens enclosure) as a right part of left-associative application for both Path and Pi lambdas.
Universe indicies $U_j$ (inner fibrant), $V_k$ (outer pretypes) and S (outer strict omega) are using unicode subscript letters that are already processed in lexer.

\begin{table}[ht!]
\tabstyle
\begin{tabular}{llll}
\textbf{menhir} \\
\multicolumn{3}{l}{\ \ \ exp2 : exp2 exp3 | exp2 APPFORMULA exp3 | exp3 } \\
\multicolumn{4}{l}{\ \ \ exp3 : LPARENS exp0 RPARENS LSQ exp0 MAP exp0 RSQ } \\
\ \ \ \ \ \   | HOLE              & | PRE          & | KAN          & | IDJ exp3 \\
\ \ \ \ \ \   | exp3 FST          & | exp3 SND     & | NEGATE exp3  & | INC exp3 \\
\ \ \ \ \ \   | exp3 AND exp3     & | exp3 OR exp3 & | ID exp3      & | REF exp3\\
\ \ \ \ \ \   | OUC exp3          & | PATHP exp3   & | PARTIAL exp3 & | IDENT \\
\multicolumn{3}{l}{\ \ \ \ \ \    | IDENT LSQ exp0 MAP exp0 RSQ }   & | HCOMP exp3 \\
\multicolumn{3}{l}{\ \ \ \ \ \    | LPARENS exp0 RPARENS }          & | TRANSP exp3 exp3 \\
\end{tabular}
\end{table}


\newpage
\section{Semantics}
The idea is to have a unified layered type checker, so you can disbale/enable any MLTT-style inference, assign types to universes and enable/disable hierachies. This will be done by providing linking API for pluggable presheaf modules. We selected 5 levels of type checker awareness from universes and pure type systems up to synthetic language of homotopy type theory. Each layer corresponds to its presheaves with separate configuration for universe hierarchies.
\begin{table}[ht]
\tabstyle
\begin{tabular}{rcl}
 \textbf{inductive} lang : \textbf{U} & := & UNI: cosmos $\rightarrow$ lang \\
   & | & PI: pure lang $\rightarrow$ lang \\
   & | & SIGMA: total lang $\rightarrow$ lang \\
   & | & ID: uip lang $\rightarrow lang$ \\
   & | & PATH: homotopy lang $\rightarrow$ lang \\
   & | & GLUE: gluening lang $\rightarrow$ lang \\
   & | & HIT: hit lang $\rightarrow$ lang \\
\end{tabular}
\end{table}
We want to mention here with homage to its authors all categorical models of dependent type theory: Comprehension Categories (Grothendieck, Jacobs), LCCC (Seely), D-Categories and CwA (Cartmell), CwF (Dybjer), C-Systems (Voevodsky), Natural Models (Awodey). While we can build some transports between them, we leave this excercise for our mathematical components library.
We will use here the Coquand's notation for Presheaf Type Theories in terms of restriction maps.

\subsection{Universe Hierarchies}

Language supports Agda-style hierarchy of universes: prop, fibrant (U), interval pretypes (V) and strict omega
with explicit level manipulation. All universes are bounded with preorder
\begin{equation}
Prop_i \prec Fibrant_j \prec Pretypes_k \prec Strict_l,
\end{equation}
in which $i,j,k,l$ are bounded with equation:
\begin{equation}
i < j < k < l.
\end{equation}
Large elimination to upper universes is prohibited. This is extendable to Agda model:

\begin{table}[ht]
\tabstyle
\begin{tabular}{rcl}
  \textbf{inductive} cosmos : \textbf{U}& := & prop: nat $\rightarrow$ cosmos \\
   &|& fibrant: nat \\
   &|& pretypes: nat \\
   &|& strict: nat \\
   &|& omega \\
   &|& lock \\
\end{tabular}
\end{table}

The \textbf{anders} model contains only fibrant $U_j$ and pretypes $V_k$ universe hierarchies.

\newpage

\subsection{Dependent Types}

\begin{definition}[Type]
A type is interpreted as a presheaf $A$, a family of sets $A_I$ with restriction maps
$u \mapsto u\ f, A_I \rightarrow A_J$ for $f: J\rightarrow I$. A dependent type
B on A is interpreted by a presheaf on category of elements of $A$: the objects
are pairs $(I,u)$ with $u : A_I$ and morphisms $f: (J,v) \rightarrow (I,u)$ are
maps $f : J \rightarrow$ such that $v = u\ f$. A dependent type B is thus given
by a family of sets $B(I,u)$ and restriction maps $B(I,u) \rightarrow B(J,u\ f)$.
\end{definition}


We think of $A$ as a type and $B$ as a family of presheves $B(x)$ varying $x:A$.
The operation $\Pi(x:A)B(x)$ generalizes the semantics of
implication in a Kripke model.

\begin{definition}[Pi]
An element $w:[\Pi(x:A)B(x)](I)$ is a family of functions $w_f : \Pi(u:A(J))B(J,u)$
for $f : J \rightarrow I$ such that $(w_f u)g=w_{f\ g}(u\ g)$ when $u:A(J)$ and $g:K\rightarrow J$.
\end{definition}
\begin{table}[ht]
\tabstyle
\begin{tabular}{rcl}
  \textbf{inductive} pure (lang : U) : \textbf{U}& := & pi: name $\rightarrow$ nat $\rightarrow$ lang $\rightarrow$ lang $\rightarrow$ pure lang \\
  &|& lambda: name $\rightarrow$ nat $\rightarrow$ lang $\rightarrow$ lang \\
  &|& app: lang $\rightarrow$ lang \\
\end{tabular}
\end{table}

\begin{definition}[Sigma]
The set $\Sigma(x:A)B(x)$ is the set of pairs $(u,v)$ when $u:A(I),v:B(I,u)$ and restriction map $(u,v)\ f=(u\ f,v\ f)$.
\end{definition}
\begin{table}[ht]
\tabstyle
\begin{tabular}{rcl}
  \textbf{inductive} total (lang : U) : \textbf{U}& := & sigma: name $\rightarrow$ lang $\rightarrow$ total lang \\
  &|& pair: lang $\rightarrow$ lang \\
  &|& fst: lang \\
  &|& snd: lang \\
\end{tabular}
\end{table}

The presheaf with only Pi and Sigma is called \textbf{MLTT-72}. Its internalization in \textbf{anders} is as follows:

\begin{table}[ht!]
\tabstyle
\begin{tabular}{l}
\textbf{def} MLTT-72 (A : U) (B : A $\rightarrow$ U) : U := $\Sigma$ \\
\ \ ($\Pi$-form$_1$ : U) \\
\ \ ($\Pi$-ctor$_1$ : Pi A B $\rightarrow$ Pi A B) \\
\ \ ($\Pi$-elim$_1$ : Pi A B $\rightarrow$ Pi A B) \\
\ \ ($\Pi$-comp$_1$ : (a : A) (f : Pi A B), $\Pi$-elim$_1$ ($\Pi$-ctor$_1$ f) a = f a) \\
\ \ ($\Pi$-comp$_2$ : (a : A) (f : Pi A B), f = $\lambda$ (x : A), f x) \\
\ \ ($\Sigma$-form$_1$ : U) \\
\ \ ($\Sigma$-ctor$_1$ : $\Pi$ (a : A) (b : B a) , Sigma A B) \\
\ \ ($\Sigma$-elim$_1$ : $\Pi$ (p : Sigma A B), A) \\
\ \ ($\Sigma$-elim$_2$ : $\Pi$ (p : Sigma A B), B (pr$_1$ A B p)) \\
\ \ ($\Sigma$-comp$_1$ : $\Pi$ (a : A) (b: B a), a = $\Sigma$-elim$_1$ ($\Sigma$-ctor$_1$ a b)) \\
\ \ ($\Sigma$-comp$_2$ : $\Pi$ (a : A) (b: B a), b = $\Sigma$-elim$_2$ (a, b)) \\
\ \ ($\Sigma$-comp$_3$ : $\Pi$ (p : Sigma A B), p = (pr$_1$ A B p, pr$_2$ A B p)), \textbf{1}
\end{tabular}
\end{table}

\newpage
\subsection{Path Equality}

The fundamental development of equality inside MLTT provers led us to the
notion of $\infty$-groupoid as spaces. In this way Path identity type appeared
in the core of type checker along with De Morgan algebra on built-in interval type.

\begin{table}[ht]
\tabstyle
\begin{tabular}{rcl}
  \textbf{inductive} homotopy (lang : U) : U & := & PathP: lang $\rightarrow$ lang $\rightarrow$ lang \\
  &|& plam: name $\rightarrow$ lang $\rightarrow$ lang \\
  &|& papp: lang $\rightarrow$ lang \\
  &|& I \\
  &|& zero \\
  &|& one \\
  &|& meet: lang $\rightarrow$ lang \\
  &|& join: lang $\rightarrow$ lang \\
  &|& neg: lang \\
  &|& system: lang \\
  &|& Partial: lang \\
  &|& transp: lang $\rightarrow$ lang \\
  &|& hcomp: lang \\
  &|& Sub: lang \\
  &|& inc: lang \\
  &|& ouc: lang \\
\end{tabular}
\end{table}

$\textbf{Definition}$ (Cubical Presheaf $\mathbb{I}$).
The identity types modeled with another presheaf, the presheaf on Lawvere
category of distributive lattices (theory of De Morgan algebras) denoted
with $\Box$ — $\textbf{I} : \Box^{op} \rightarrow \textrm{Set}$.

$\textbf{Definition}$ Properties of $\textbf{I}$. The presheaf $\textbf{I}$:
i) has to distinct global elements $0$ and $1$ (B$_1$);
ii) $\textbf{I}$(I) has a decidable equality for each $I$ (B$_2$);
iii) $\textbf{I}$ is tiny so the path functor $X \mapsto X^\textbf{I}$ has right adjoint (B$_3$).;
iv) $\textbf{I}$ has meet and join (connections).\\

$\textbf{Interval\ Pretypes}$. While having pretypes universe V with interval and
associated De Morgan algebra ($\wedge$, $\vee$, $-$, $0$, $1$, $\textrm{I}$) is enough to
perform DNF normalization and proving some basic statements about path, including:
contractability of singletons, homotopy transport, congruence, functional
extensionality; it is not enough for proving $\beta$ rule for Path type or path composition.\\

$\textbf{Generalized\ Transport}$. Generalized transport transp adresses
first problem of deriving the computational $\beta$ rule for Path types:

\begin{table}[ht]
\tabstyle
\begin{tabular}{l}
\textbf{theorem} Path-$\beta$ (A : U) (a : A) (C : D A) (d: C a a (refl A a)) \\
\ \ \ \ : Equ (C a a (refl A a)) d (J A a C d a (refl A a)) \\
\ \ \ \ := $\lambda$ (A : U), $\lambda$ (a : A), $\lambda$ (C : $\Pi$ (x : A), $\Pi$ (y : A), PathP (<\_> A) x y $\rightarrow$ U), \\
\ \ \ \ \ \ \ $\lambda$ (d : C a a (<\_> a)), <j> transp (<\_> C a a (<\_> a)) -j d \\
\end{tabular}
\end{table}

Transport is defined on fibrant types (only) and type checker should cover all the cases
Note that transp$^i$ (Path$^j$ A v w) $\varphi$ u$_0$ case is relying on comp operation which depends on hcomp primitive.

\begin{table}[ht]
\tabstyle
\begin{tabular}{l}
transp$^{i}$ N $\varphi$ u$_0$ = u$_0$ \\
transp$^{i}$ U $\varphi$ A = A \\
transp$^{i}$ ($\Pi$ (x : A), B) $\varphi$ u$_0$ v = transp$^i$ B(x/w) $\varphi$ (u$_0$ w(i/0)) \\
transp$^{i}$ ($\Sigma$ (x : A), B) $\varphi$ u$_0$ = (transp$^i$ A $\varphi$ (u$_0$.1),transp$^i$ B(x/v) $\varphi$ (u$_0$.2)) \\
transp$^{i}$ (Path$^j$ v w) $varphi$ u$_0$ = <j> comp$^i$ A [$\phi$ u$_0$ j, (j=0) $\mapsto$ v, (j=1) $\mapsto$ w] (u$_0$ j) \\
transp$^{i}$ (Glue [$\varphi$ $\mapsto$ (T,w)] A) $\psi$ u$_0$ = glue [$\phi$(i/1) $\mapsto$ t'$_1$] a'$_1$ : B(i/1) \\
\end{tabular}
\end{table}

\newpage
$\textbf{Partial\ Elements}$. In order to explicitly define hcomp we need to specify
n-cubes where some faces are missing. Partial primitives isOne, 1=1 and UIP on pretypes
are derivable in Anders due to landing strict equality Id in V universe. The idea is
that (Partial A r) is the type of cubes in A that are only defined when IsOne r holds.
(Partial A r) is a special version of the function space IsOne r → A with a more
extensional equality: two of its elements are considered judgmentally equal if
they represent the same subcube of A. They are equal whenever they reduce to
equal terms for all the possible assignment of variables that make r equal to 1.

\begin{table}[ht]
\tabstyle
\begin{tabular}{l}
def Partial' (A : U) (i : I) := Partial A i \\
def isOne : I -> V := Id I 1 \\
def 1=>1 : isOne 1 := ref 1 \\
def UIP (A : V) (a b : A) (p q : Id A a b) : Id (Id A a b) p q := ref p \\
\end{tabular}
\end{table}

$\textbf{Cubical\ Subtypes}$. For (A : U) (i : I) (Partial A i) we can define
subtype A [ i $\mapsto$ u ]. A term of this type is a term of type A that is
definitionally equal to u when (IsOne i) is satisfied. We have forth and back
fusion rules ouc (inc v) = v and inc (outc v) = v. Moreover, ouc v will reduce to u 1=1 when i=1.

\begin{table}[ht]
\tabstyle
\begin{tabular}{l}
def sub' (A : U) (i : I) (u : Partial A i) : V := A [i $\mapsto$ u ] \\
def inc' (A : U) (i : I) (a : A) : A [i $\mapsto$ [(i = 1) $\rightarrow$ a]] := inc A i a \\
def ouc' (A : U) (i : I) (u : Partial A i) (a : A [i $\mapsto$ u]) : A := ouc a \\
\end{tabular}
\end{table}

$\textbf{Homogeneous\ Composition}$. hcomp is the answer to second problem: with hcomp and transp
one can express path composition, groupoid, category of groupoids (groupoid interpretation and
internalization in type theory). One of the main roles of homogeneous composition is to be a
carrier in [higher] inductive type constructors for calculating of homotopy colimits and direct
encoding of CW-complexes (or internal models as gluening of pushout cells).

\begin{table}[ht]
\tabstyle
\begin{tabular}{l}
hcomp$^i$ N [$\phi$ $\mapsto$ 0] 0 = 0 \\
hcomp$^i$ N [$\phi$ $\mapsto$ S u] (S u$_0$) = S (hcomp$^i$ N [$\phi$ $\mapsto$ u] u$_0$) \\
hcomp$^i$ U [$\phi$ $\mapsto$ E] A = Glue [$\phi$ $\mapsto$ (E(i/1), equiv$^i$ E(i/1-i))] A \\
hcomp$^i$ ($\Pi$ (x : A), B) [$\phi$ $\mapsto$ u] u$_0$ v = hcomp$^i$ B(x/v) [$\phi$ $\mapsto$ u v] (u$_0$ v) \\
hcomp$^i$ ($\Sigma$ (x : A), B) [$\phi$ $\mapsto$ u] u$_0$ = (v(i/1), comp$^i$ B(x/v) [$\phi$ $\mapsto$ u.2] u$_0$.2) \\
hcomp$^i$ (Path$^j$ A v w) [$\phi$ $\mapsto$ u] u$_0$ = <j> hcomp$^i$ A [ $\phi$ $\mapsto$ u j, (j = 0) $\mapsto$ v, (j = 1) $\mapsto$ w ] (u$_0$ j) \\
hcomp$^i$ (Glue [$\phi$ $\mapsto$ (T,w)] A) [$\psi$ $\mapsto$ u] u$_0$ \\
\ \ \ \ = glue [$\phi$ $\mapsto$ t$_1$] a$_1$ \\
\ \ \ \ = glue [$\phi$ $\mapsto$ u(i/1)] (unglue u(i/1)) \\
\ \ \ \ = u(i/1) : Glue [$\phi$ $\mapsto$ (T,w)] A \\

hfill$^i$ A [$\phi$ $\mapsto$ u] u$_0$ = hcomp$^j$ A [$\phi$ $\mapsto$ u(i/i$\vee$j), (i=0) $\mapsto$ u$_0$] u$_0$ : A \\
v = hfill$^i$ A [$\phi$ $\mapsto$ u.1] u$_0$.1 \\
t$_1$ = u(i/1) : T \\
a$_1$ = unglue u(i/1) : A \\
glue [$\phi$ $\mapsto$ t$_1$] a1 = t$_1$ : T \\
\end{tabular}
\end{table}

\subsection{Strict Equality}

To avoid conflicts with path equalities which live in fibrant universes strict
equalities live in pretypes universes.

\begin{table}[ht]
\tabstyle
\begin{tabular}{rcl}
  \textbf{inductive} strict (lang : U) : U & := & Id: name $\rightarrow$ lang \\
  &|& ref: lang $\rightarrow$ lang \\
  &|& idJ: lang $\rightarrow$ lang $\rightarrow$ lang \\
\end{tabular}
\end{table}

We use strict equality in HTS for pretypes and partial elements which live in V.
The presheaf configuration with Pi, Sigma and Id is called MLTT-73. The presheaf
configuration with Pi, Sigma, Id and Path is called HTS.

\newpage
\subsection{Glue Types}

The main purpose of Glue types is to construct a cube where some faces have
been replaced by equivalent types. This is analogous to how hcomp lets us
replace some faces of a cube by composing it with other cubes, but for Glue
types you can compose with equivalences instead of paths. This implies the
univalence principle and it is what lets us transport along paths built
out of equivalences.

\begin{table}[ht]
\tabstyle
\begin{tabular}{rcl}
  \textbf{inductive} glue (lang : U) : U & := & Glue: lang $\rightarrow$ lang $\rightarrow$ lang \\
  &|& glue: lang $\rightarrow$ lang \\
  &|& unglue: lang $\rightarrow$ lang \\
\end{tabular}
\end{table}

\subsection{Higher Inductive Types}

Anders currently don’t support Lean-compatible generic inductive schemes
definition. So instead of generic inductive schemes Anders supports well-founded trees (W-types).
Basic data types like List, Nat, Fin, Vec are implemented as W-types in base library.
As for higher inductive types Anders has Three-HIT basis (Coequalizer, HubSpoke and Colimit) to express other HIT.

\subsection{de Rham Stack}

Stack de Rham or Infinitezemal Shape Modality is a basic primitive for proving theorems
from synthetic differential geometry. This type-theoretical framework was developed
for the first time by Felix Cherubini under the guidance of Urs Schreiber. The Anders
prover implements the computational semantics of the de Rham stack.

\bibliography{anders}

\end{document}
